---
title: "Homework 7"
author: "Lee Steffen"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE,
                      warning = FALSE, error = TRUE,
                      fig.height = 3)
library(tidyverse)
```

# Preliminaries

- This file should be in `STAT240/homework/hw07` on your local computer.


# Problem 1

Let $X_1$ and $X_2$ be two draws from $X \sim N(10, 4)$. Order the five events below based on which events are least to most likely to occur.

- Event A: $X_1 > 15$
- Event B: $X_1 = 15$
- Event C: $X_1 < 15$
- Event D: $X_1 > 15$ AND $X_2 > 15$
- Event E: $X_1 > X_2$

Least --> Most Likely
- Event B
- Event D
- Event A
- Event E
- Event C

# Problem 2

The weight of adult male grizzly bears in the continental United States is well approximated by a normal distribution with mean 510 pounds and standard deviation 45 pounds. The weight of adult female grizzly bears in the continental United States is well approximated by a normal distribution with mean 315 pounds and standard deviation 37 pounds.

Suppose a male grizzly bear that is 441 pounds is observed. What would be the approximate weight of a female grizzly bear with the same weight percentile as this male grizzly bear?

```{r}
input <- pnorm(441, 510, 45)

approx_weight <- qnorm(input, 315, 37)

approx_weight
```

# Problem 3

The lifespan of a certain type of tire is approximately normal, with mean 39 and standard deviation 2.5 (units are in thousands of miles).
  
(a) What is the probability that a random tire will last longer than 40,000 miles?

```{r}
pnorm(40, 39, 2.5, lower.tail = F)
```
(b) What is the 95th percentile of tire lifespan?

```{r}
qnorm(0.95, 39, 2.5)
```

(c) Take a random sample of 4 tires.  What is the probability that the average lifespan of the four tires will be greater than 40,000 miles?

```{r}
pnorm(40, 39, 2.5 / sqrt(4), lower.tail = F)
```
(d) What is the 95th percentile for the average lifespan of four tires?  

```{r}
qnorm(0.95, 39, 2.5 / sqrt(4))
```

(e) Explain why the answers to (a) and (b) are so different from the ansewrs to (c) and (d).

In problems (a) and (b) we're talking about only a single tire, but in problems (c) and (d) we're talking about 4 tired which changes the probabilities when considering 4 instead of just a single tire.

# Problem 4

Suppose you are playing a coin flipping game with a friend, where you suspect the coin your friend provided is not a fair coin.  In fact, you think the probability the coin lands heads is less than 0.5.  To test this, you flip the coin 100 times and observe the coin lands heads 30 times.
  
(a) If you assume the coin is fair (i.e., the probability of the coin landing heads is 0.5), what is the probability of observing 30 heads or fewer, calculated using an exact model?

```{r}
pbinom(30, 100, 0.5)
```
(b) Calculate the previous probability, but use a normal approximation to achieve a numerical value. What is the error in this approximation?

```{r}
mu <- 100 * 0.5
sd <- sqrt(mu*(1-0.5))

pnorm(30.5, mu, sd)
```

(c) How small would $p$ need to be (rounded to the nearest 0.01) for the probability of observing 30 or fewer heads to be greater than 0.05? *You could determine $p$ by trial and error, but there is also a quicker solution using `tibble()`.*

```{r}

tibble(p = seq(0.01, 0.50, by = 0.01)) %>%
  mutate(prob = pbinom(30, 100, p)) %>%
  filter(prob > 0.05)
```
Based on this code we see that the smallest p would have to be 0.38

# Problem 5

The code below does the following: 

- Draw `n` points from a population called `pop`
- Calculate the mean of the `n` points
- Repeat this 10000 times to get a large number of sample means

```{r, eval = FALSE}
replicate(10000, mean(sample(pop, n, replace = T)))
```

Copy-paste and edit the line above to perform this operation three times with 10000 iterations each.  Do this with $n = 5$, $n = 10$, and $n = 50$. Use `right_skewed` below as your population, `pop`.

Then, make four histograms to compare the original population and the three sampling distributions.  What do you notice as $n$ increases?

```{r}
right_skewed <- c(rep(0, 60), rep(1, 25), rep(2, 10), rep(3, 5))

means_5 <- replicate(10000, mean(sample(right_skewed, 5, replace = T)))
means_10 <- replicate(10000, mean(sample(right_skewed, 10, replace = T)))
means_50 <- replicate(10000, mean(sample(right_skewed, 50, replace = T)))

hist(right_skewed, main = "Original Population", xlab = "Value")
hist(means_5, main = "Sample Means (n = 5)", xlab = "Sample Mean")
hist(means_10, main = "Sample Means (n = 10)", xlab = "Sample Mean")
hist(means_50, main = "Sample Means (n = 50)", xlab = "Sample Mean")
```

# Problem 6

A 95% confidence interval for the true mean monthly rent in USD in Madison based on a random sample of 100 apartments is (943, 1024).  Briefly explain whether each of the following statements are true or false.

(a) About 95 percent of all rental prices in Madison are between 943 and 1024.

False, it's more about confidence rather than an answer to how much there is

(b) About 95 percent of the 100 rental prices in this sample are between 943 and 1024.

False, a confidence interval does not describe where 95% of the sample data fall. It describes plausible values for the true mean

(c) In repeated sampling, the method produces intervals that include the population mean in approximately 95% of samples. 

True, in repeated random samples, about 95% of confidence intervals constructed
this way would contain the true population mean

(d) There is a 95% chance that the true mean monthly rent is between 943 and 1024.

False, after the interval is calculated, the true mean is fixed. The interval either contains it or it does not. The 95% refers to the long run success rate of the method, not the probability for this one interval

(e) Working with this same sample, we would expect a hypothesis test with significance level 0.05 to fail to reject a null hypothesis that the true mean monthly rent is 1000.

True, we can conduct a hypothesis test and check if the significance level rejects the null hypothesis

# Problem 7

For each of the three prompts, explain which condition will result in a *narrower* confidence interval.  If the confidence level would not change, explain why.

- Having a larger confidence level (smaller alpha) versus having a smaller confidence level (larger alpha)

Narrower confidence interval we be having a smaller confidence level (larger alpha)

- Having larger estimation error (a.k.a "sampling error") versus having smaller estimation error

A confidence interval becomes narrower when the estimation error decreases because less is added and subtracted from the estimate.

- Having a larger point estimate versus having a smaller point estimate

Confidence level doesn't change because it's just there to help us calculate the ranges and only changes the center

# Problem 8

The code below calculates the Z critical value to build a 95% confidence interval.

```{r}
qnorm(0.975)
```

(a) Edit the code to instead calculate the critical value for 98% confidence.

```{r}
qnorm(0.99)
```

(b) Next, generalize this operation to any confidence level.  Uncomment and edit the code below to return the critical value for a (1-alpha) confidence interval for any choice of alpha.

```{r}
# For example, a 90% CI
alpha <- 0.1

qnorm(0.95)
```

(c) What would be the critical value for a 100% confidence interval?

```{r}
qnorm(0)
```

